---
title: "Reproducible Research: Peer Assessment 1"
output: 
    html_document:
        keep_md: true
        fig_width: 9
        fig_height: 6
        toc: true
---


## Loading and preprocessing the data

*Show any code that is needed to:*

1. *Load the data (i.e. `read.csv()`).*
2. *Process/transform the data (if necessary) into a format suitable for your analysis.*

```{r, message = FALSE}
library(xts)
library(lubridate)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
# library(devtools)
# install_github("jrnold/ggthemes")
library(ggthemes)
library(pander)

options(scipen = 1, digits = 2)
```

The data is provided as a CSV file `activity.csv`. First, it is loaded as a `tbl_df` object, suitable for further processing with `dplyr`. Next, dates and intervals are transformed into a POSIX compliant form:

a) `hour` is a POSIX time variable, coupled with a dummy date (1900-01-01),
b) `time` is a POSIX date-time variable, indicating 5-minute intervals (eg. 2012-10-01 00:05).

```{r}
activity <- "data/activity.csv" %>%
    read.csv(stringsAsFactors = FALSE) %>%
    tbl_df %>%
    mutate(
        hour.string = sprintf("%04d", interval),
        hour = ymd_hm(paste("1900-01-01", hour.string)),
        time = ymd_hm(paste(date, hour.string))
    )
```

Radom 10 rows are presented below.

```{r}
activity %>% sample_n(10) %>% arrange(time) %>% pander(split.table = Inf)
```

## What is mean total number of steps taken per day?

*For this part of the assignment, you can ignore the missing values in the dataset.*

1. *Make a histogram of the total number of steps taken each day.*
2. *Calculate and report the mean and median total number of steps taken per day.*

In this step, data is aggregated in daily intervals.

```{r}
activity.daily <- activity %>%
    group_by(date) %>%
    summarise(steps = sum(steps))

activity.daily %>% sample_n(10) %>% arrange(date) %>% pander(split.table = Inf)
```

Histogram of daily aggregated number of steps.

```{r}
activity.daily %>%
    ggplot +
    geom_histogram(aes(steps), binwidth = 500) +
    theme_pander() +
    scale_colour_pander()
```

Summary of `steps` variable.

```{r}
activity.stats <- activity.daily$steps %>%
    summary(digits = 5) %>%
    data_frame(stat = names(.), steps = .)

activity.stats %>% pander

# activity.daily$steps %>% mean(na.rm = TRUE)
# activity.daily$steps %>% median(na.rm = TRUE)
```

1. The **mean** is `r activity.daily$steps %>% mean(na.rm = TRUE)`.
2. The **median** is `r activity.daily$steps %>% median(na.rm = TRUE)`.

## What is the average daily activity pattern?

1. *Make a time series plot (i.e. `type = "l"`) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).*
2. *Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?*

In this step, data is averaged across all days.

```{r}
activity.hourly <- activity %>%
    group_by(interval, hour) %>%
    summarise(steps = mean(steps, na.rm = TRUE)) %>%
    ungroup

activity.hourly %>% sample_n(10) %>% arrange(hour) %>% pander(split.table = Inf)
```

Time series plot.

```{r}
activity.hourly %>%
    ggplot +
    geom_line(aes(hour, steps)) +
    scale_x_datetime(labels = date_format("%H:%M")) +
    theme_pander() +
    scale_colour_pander()

activity.hourly %>%
    filter(steps == max(steps))
```

The maximum number of steps (`r activity.hourly %>% filter(steps == max(steps)) %>% .$steps`) happens at the interval of `r activity.hourly %>% filter(steps == max(steps)) %>% .$hour %>% format("%H:%M")`.

## Imputing missing values

*Note that there are a number of days/intervals where there are missing values (coded as `NA`). The presence of missing days may introduce bias into some calculations or summaries of the data.*

1. *Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with `NA`s).*
2. *Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.*
3. *Create a new dataset that is equal to the original dataset but with the missing data filled in.*
4. *Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?*

Count of complete cases (*i.e.* with no `NA`s) in the data set.

```{r}
activity %>% complete.cases %>% table
```

The total number of rows with missing values is `r activity %>% complete.cases %>% table %>% .["FALSE"]`.

We are going to impute missing values for any interval by substituting it with the average value for that interval, averaged across all days.

```{r}
activity <- activity %>%
    left_join(activity.hourly %>% select(interval, steps.imputed = steps), by = "interval") %>%
    mutate(steps.imputed = ifelse(is.na(steps), round(steps.imputed), steps))
```

Again, data is aggregated in daily intervals.

```{r}
activity.daily <- activity %>%
    group_by(date) %>%
    summarise(steps = sum(steps), steps.imputed = sum(steps.imputed))

activity.daily %>% sample_n(10) %>% arrange(date) %>% pander(split.table = Inf)
```

Histogram of daily aggregated number of steps.

```{r}
activity.daily %>%
    ggplot +
    geom_histogram(aes(steps.imputed), binwidth = 500) +
    theme_pander() +
    scale_colour_pander()
```

Summary of `steps.imputed` variable, compared to `steps`.

```{r}
activity.stats <- activity.stats %>%
    left_join(
        activity.daily$steps.imputed %>%
            summary(digits = 5) %>%
            data_frame(stat = names(.), steps.imputed = .),
        by = "stat"
    )

activity.stats %>% pander

# activity.daily$steps.imputed %>% mean(na.rm = TRUE)
# activity.daily$steps.imputed %>% median(na.rm = TRUE)
```

1. The **mean** is `r mean(activity.daily$steps.imputed, na.rm = TRUE)`, which is almost the same as the mean of the original data (`r mean(activity.daily$steps, na.rm = TRUE)`).
2. The **median** is `r median(activity.daily$steps.imputed, na.rm = TRUE)`, which is slightly less than the median of the original data (`r median(activity.daily$steps, na.rm = TRUE)`).

## Are there differences in activity patterns between weekdays and weekends?

*For this part the `weekdays()` function may be of some help here. Use the dataset with the filled-in missing values for this part.*

1. *Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.*
2. *Make a panel plot containing a time series plot (i.e. `type = "l"`) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).*

Adding a factor variable `day.type`.

```{r}
activity <- activity %>%
    mutate(
        day.type = factor(
            wday(time) %in% c(1, 7),
            labels = c("weekday", "weekend")
        )
    )
```

In this step, data is averaged across all days, but in two groups: weekdays and weekends.

```{r}
activity.hourly.weekend <- activity %>%
    group_by(day.type, hour) %>%
    summarise(steps.imputed = mean(steps.imputed, na.rm = TRUE)) %>%
    ungroup

activity.hourly.weekend %>% sample_n(10) %>% arrange(day.type, hour) %>% pander(split.table = Inf)
```

Time series plot.

```{r}
activity.hourly.weekend %>%
    ggplot +
    geom_line(aes(hour, steps.imputed)) +
    facet_grid(day.type ~ .) +
    scale_x_datetime(labels = date_format("%H:%M")) +
    theme_pander() +
    scale_colour_pander()
```

## Session Info

```{r}
sessionInfo()
```
